A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-05 tex/codatadp.tex
more

@@

\begin{abstract}
Codatatypes naturally capture potentially infinite data structures and
processes. We introduce a decision procedure that combines reasoning about
datatypes and codatatypes. The dual of the cyclicity rule for datatypes is a
bisimilarity rule that identifies observationally equal, possibly cyclic values.
The procedure is complete for ground problems and is compatible with the
Nelson--Oppen method. It has been implemented in the latest version of CVC4, a
modern SMT solver. An evaluation based on problems generated from Isabelle
theories demonstrates the potential of the procedure for both proving and model
finding.
\end{abstract}

@@

Freely generated datatypes are ubiquitous in functional programs and
logical specifications. They can be implemented efficiently and enjoy
properties that can be exploited in automated reasoners. They are especially
useful to represent finite data structures in computer science applications but
also arise in mathematical formalizations.
%
%However, because datatype values correspond to finite ground terms, they
%are generally not adequate to represent infinite objects.
%For example, the datatype of natural numbers
%constructed by $\const{Z} : \ty{nat}$ and $\const{S} : \ty{nat} \to \ty{nat}$
%only allow values of the form $\const{S}(\ldots(\const{S}(\const{Z}))\ldots)$.

To represent infinite values, % such as $\const{S}(\const{S}(\const{S}(\ldots))$,
a natural choice is to turn to codatatypes (or coinductive datatypes), the
non-well-founded dual of (inductive) datatypes.
%
Although they are sometimes perceived as esoteric, codatatypes have a role to
play in computer science. For example, the verified C compiler CompCert
\cite{leroy-2009}, the verified Java compiler JinjaThreads
\cite{lochbihler-2010-jinja}, and the formalized Java memory model
\cite{lochbihler-2014-jmm} all rely crucially on codatatypes to capture infinite
processes.
% as do the CoALP dialect of Prolog and the Dafny specification language.

Codatatypes, like datatypes, are freely generated by their constructors;
but for them, infinite constructor terms are also legitimate values
(Section~\ref{sec:theory-of-co-datatypes}). Intuitively, a
codatatype consists of all finite and infinite ground constructor terms, and
only those. For example, the specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that aptly models the natural numbers extended with an
infinite value $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$. Compared
with the more conventional, inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Inf}\]
%
the codatatype eliminates one case by unifying the finite and infinite cases.
Equations such as $\const{S}(\infty) \teq \infty$ hold without any additional
work, because both sides share the same infinite expansion
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies a
codatatype value.

@@

Datatypes and codatatypes share many of the same properties, so it makes sense
to consider them together. The main differences are as follows:
%
\begin{itemize}
\item
No well-foundedness check; e.g. the type of infinite sequences, or streams, over an
element type $\tau$
%
\[\keyw{codatatype}~\;\ty{stream}_{\,\tau} \,=\, \const{SCons}(\tau,\, \ty{stream}_{\,\tau})\]
is allowed, whereas the corresponding datatype would be rejected, because no
finite ground constructor terms exist.

\cheat
\item
The cyclicity check must be dropped. A constraint such as
$\const{x} \teq \const{C}(\const{x})$, where $\const{C}$ is a constructor,
is unsatisfiable for datatypes but satisfiable for codatatypes.
Instead: bisimilarity check, which states that two values having the same (possibly
infinite) expansion must be equal. From $\const{x} \teq \const{C}(\const{y})$ and
$\const{y} \teq \const{C}(\const{x})$, deduce $\const{x} = \const{y}$.

\cheat
\item
Normal forms are needed and must be enumeratable. For finite model finding, requires a $\mu$ notation,
e.g., $\const{SCons}(0,\, \mu s.\; \const{SCons}(1,\, \const{SCons}(2,\, s)))$
represents the lasso-shaped sequence $0, 1, 2, 1, 2, \ldots$.
\end{itemize}

@@

The procedure is implemented in CVC4 as rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}). It implements a precise strategy for applying
inferences. It consists of about 2000 lines of \cpp{} code, among which 1600 are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the 1.5 release.
%
The decision procedure is evaluated on %hand-crafted examples and on
problems generated from Isabelle theories using Isabelle's Sledgehammer tool
(Section~\ref{sec:experimental-results}).

>>>>

\let\labelitemi=\labelitemii %% CHEAT!

@@

\begin{abstract}
Codatatypes naturally capture potentially infinite data structures and
processes. We present a decision procedure that combines reasoning about
datatypes and codatatypes. The dual of the cyclicity rule for datatypes is a
bisimilarity rule that identifies equal values, also in the presence of cycles.
The procedure is complete for ground problems and is compatible with the
Nelson--Oppen method. It has been implemented in the latest version of CVC4, a
modern SMT solver. An evaluation based on problems generated from Isabelle
theories demonstrates the potential of the procedure for both proving and model
finding.
\end{abstract}

@@

Freely generated datatypes are ubiquitous in functional programs and
logical specifications. They are especially
useful to represent finite data structures in computer science applications but
also arise in mathematical formalizations.
They can be implemented efficiently and enjoy
properties that can be exploited in automated reasoners. 
%
%However, because datatype values correspond to finite ground terms, they
%are generally not adequate to represent infinite objects.
%For example, the datatype of natural numbers
%constructed by $\const{Z} : \ty{nat}$ and $\const{S} : \ty{nat} \to \ty{nat}$
%only allow values of the form $\const{S}(\ldots(\const{S}(\const{Z}))\ldots)$.

To represent infinite values, % such as $\const{S}(\const{S}(\const{S}(\ldots))$,
a natural choice is to turn to codatatypes (or coinductive datatypes), the
non-well-founded dual of (inductive) datatypes.
%
Although they are sometimes perceived as esoteric, codatatypes have a role to
play in computer science. For example, the verified C compiler CompCert
\cite{leroy-2009}, the verified Java compiler Jinja\-Threads
\cite{lochbihler-2010-jinja}, and the formalized Java memory model
\cite{lochbihler-2014-jmm} all rely crucially on codatatypes to capture infinite
processes.
% as do the CoALP dialect of Prolog and the Dafny specification language.

Codatatypes are freely generated by their constructors, but unlike for datatypes
infinite constructor terms are also legitimate values
(Section~\ref{sec:theory-of-co-datatypes}). Intuitively, a
codatatype consists of all finite and infinite ground constructor terms, and
only those. For example, the specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that faithfully
models the natural numbers extended with an
infinity $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$. Compared
with the more conventional, inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Inf}\]
%
the codatatype eliminates one case by unifying the finite and infinite cases.
Moreover, equations such as $\const{S}(\infty) \teq \infty$ hold without additional
work, because both sides share the same infinite expansion
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies the
codatatype value.

@@

Datatypes and codatatypes share many of the same properties, so it makes sense
to consider them together. The main differences are as follows:
%
\begin{itemize}
\item
\emph{No well-foundedness check is necessary for codatatypes.}
For example, the type of infinite sequences, or streams, over an
element type $\tau$
%
\[\keyw{codatatype}~\;\ty{stream}_{\,\tau} \,=\, \const{SCons}(\tau,\, \ty{stream}_{\,\tau})\]
is allowed; the corresponding datatype would be rejected as
non-well-founded.

\cheat
\item
\emph{A bisimilarity rule takes the place of the cyclicity rule of datatypes.}
A cyclic constraint such as
$\const{x} \teq \const{S}(\const{x})$ %, where $\const{C}$ is a constructor,
is unsatisfiable for a datatype but satisfiable for a codatatype.
The bisimilarity principle states that two values having the same (possibly
infinite) expansion must be equal. From $\const{x} \teq \const{S}(\const{y})$ and
$\const{y} \teq \const{S}(\const{x})$, it deduces $\const{x} \teq \const{y}$.

\cheat
\item
Normal forms are needed and must be enumeratable. For finite model finding, requires a $\mu$ notation,
e.g., $\const{SCons}(3,\, \mu s.\; \const{SCons}(1,\, \const{SCons}(4,\, s)))$
represents the lasso-shaped sequence $3, 1, 4, 1, 4, \ldots$\vthinspace.
\end{itemize}

@@

The procedure is implemented in the SMT solver CVC4 as rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}). It implements a precise strategy for applying
inferences. It consists of about 2000 lines of \cpp{} code, among which 1600 are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the 1.5 release.
%
The decision procedure is evaluated on %hand-crafted examples and on
problems generated from Isabelle theories using Isabelle's Sledgehammer tool
(Section~\ref{sec:experimental-results}).
